---
title: "Week2 Project"
author: "Patrick Ayer"
date: "6/9/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
Question 1:
```{r read data}
adata <- read.csv("activity.csv")
head(adata, n = 30)
```

Question 2:
- remove the rows with missing values, as they won't be needed
- create new dataset of complete rows, stored in `ccdata` but keep original dataset

```{r complete data}
ccdata <- adata[complete.cases(adata), ]
head(ccdata, n = 30)
```

## What is mean total number of steps taken per day?
Question 1:
```{r histogram 1}
dates <- unique(as.factor(ccdata$date))
l <- length(dates)
total_steps <- vector("numeric", length = l)
for (i in 1:l)
{
    ddata <- ccdata[ccdata$date == as.character(dates[i]), ]
    total_steps[i] <- sum(ddata$steps)
}
head(total_steps, n = 30)
hist(total_steps, breaks = l, xlab = "steps", main = "Steps per day")
```
Question 2: mean and median total steps taken per day:
```{r mean/median 1}
print(paste("Mean is ", mean(total_steps)), sep = "")
print(paste("Median is ", median(total_steps)), sep = "")
```

## What is the average daily activity pattern?
Question 1:
```{r daily activity pattern}
intervals <- unique(as.factor(ccdata$interval))
averages <- vector("numeric", length = length(intervals))
max_avg <- 0
max_interval <- 0
for (i in 1:length(intervals))
{
    idata <- ccdata[ccdata$interval == as.character(intervals[i]), ]
    averages[i] <- mean(idata$steps)
    if (averages[i] > max_avg)
    {
        max_avg <- averages[i]
        max_interval <- intervals[i]
    }
}
head(averages, n = 20)
plot(intervals, averages, type = "l")
max_avg
as.integer(as.character(max_interval))
```
Question 2: the 5-minute interval with the maximum number of steps, on average, is the interval `r as.integer(as.character(max_interval))`, with a value of `r max_avg`

## Imputing missing values
Question 1: the total number of missing values is `r nrow(adata) - nrow(ccdata)`
Question 2 & 3: the strategy I will use in filling in the missing data is by replacing all NA values with the means for that time interval
```{r replace NAs}
for (n in 1:nrow(adata))
{
    if (is.na(adata[n, 1]))
    {
        this_i <- adata[n, 3]
        for (i in 1:length(intervals))
        {
            if (intervals[i] == this_i)
            {
                adata[n, 1] <- averages[i]
            }
        }
    }
}
head(adata, n = 30)
```
Question 4: histogram of total steps per day
```{r histogram 2}
dates <- unique(as.factor(adata$date))
l <- length(dates)
total_steps <- vector("numeric", length = l)
head(total_steps, n = 10)
for (i in 1:l)
{
    ddata <- adata[adata$date == as.character(dates[i]), ]
    total_steps[i] <- sum(ddata$steps)
}
head(total_steps, n = 30)
hist(total_steps, breaks = l, xlab = "steps", main = "Steps per day")
```
mean and median total number of steps taken per day:
```{r mean/median 2}
print(paste("Mean is ", mean(total_steps)), sep = "")
print(paste("Median is ", median(total_steps)), sep = "")
```
It's strange that these values are exactly the same, but they are very close to the original values with only the complete data set. The impact of estimating the missing values appears to be negligible.

## Are there differences in activity patterns between weekdays and weekends?

Question 1: below is the code chunk to create a new factor variable in the dataset to indicate weekends and weekdays:
```{r create weekdays}
wdata <- ccdata
wdata$daytypes <- ""
for (i in 1:nrow(wdata))
{
    day <- weekdays(as.Date(wdata[i, 2]))
    if (day == "Saturday" | day == "Sunday")
    {
        wdata[i, 4] <- "weekend"
    }
    else
    {
        wdata[i, 4] <- "weekday"
    }
}

dweekdays <- wdata[wdata$daytypes == "weekday", ]
dweekends <- wdata[wdata$daytypes == "weekend", ]
head(dweekdays, n = 30)
head(dweekends, n = 30)
```
Question 2: panel plot of time intervals and average steps taken. 
```{r plot weekdays}
ints <- unique(dweekdays$interval)
l <- length(ints)
avgs <- vector("numeric", length = l)
days <- rep("weekday", l)
for (i in 1:l)
{
    i_data <- dweekdays[dweekdays$interval == ints[i], ]
    avgs[i] <- mean(i_data$steps)
}
panel_data1 <- cbind(ints, avgs, days)

ints <- unique(dweekends$interval)
l <- length(ints)
avgs <- vector("numeric", length = l)
days <- rep("weekend", l)
for (i in 1:l)
{
    i_data <- dweekends[dweekends$interval == ints[i], ]
    avgs[i] <- mean(i_data$steps)
}
panel_data2 <- cbind(ints, avgs, days)

panel_data <- rbind(panel_data1, panel_data2)
panel_data <- data.frame(panel_data)
panel_data$avgs <- as.numeric(as.character(panel_data$avgs))
panel_data$ints <- as.integer(as.character(panel_data$ints))
panel_data$days <- as.factor(panel_data$days)

library(ggplot2)
sp <- ggplot(panel_data, aes(x = ints, y = avgs)) + geom_line() + facet_grid(days ~ .)
sp
```
















